---
title: "Final project analysis"

output:

  pdf_document:

    latex_engine: xelatex

author: "Wen Du"

date: "2026-04-27"
---


```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)

```

1. Load Data

```{r}
df <- read_csv("../data/data_final_project.csv")

glimpse(df)
```

2. Create Pedon-level Dataset
```{r}
pedon_df <- df %>%
  filter(pedon_start == TRUE) %>%
  mutate(
    order_group = case_when(
      str_detect(order, regex("Histosol|Folisol|Mesisol|Humisol|Cryo", ignore_case = TRUE)) ~ "Organic soils",
      str_detect(order, regex("Podzol", ignore_case = TRUE)) ~ "Podzol",
      str_detect(order, regex("Brunisol|Brunisolic", ignore_case = TRUE)) ~ "Brunisol",
      str_detect(order, regex("Gleysol|Gleysolic", ignore_case = TRUE)) ~ "Gleysol",
      str_detect(order, regex("Regosol|Regosolic", ignore_case = TRUE)) ~ "Regosol",
      TRUE ~ "Other"
    )
  ) %>%
  select(pedon_id, order, order_group, total_c_1m, lat) %>%
  filter(!is.na(total_c_1m), !is.na(order_group))

# check structure
glimpse(pedon_df)

# check sample sizes
pedon_df %>% count(order)


```


3. Exploratory Analysis

Boxplot: Soil Carbon by Soil Order
```{r}
ggplot(pedon_df, aes(x = order_group, y = total_c_1m, fill = order_group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.1) +
  theme_classic() +
  labs(
    title = "Soil Carbon by Soil Group",
    x = "Soil Group",
    y = "Total Carbon (0-1 m)"
  )
```


4. ANOVA: Does Carbon Differ Across Soil Orders?

```{r}
aov_model <- aov(total_c_1m ~ order_group, data = pedon_df)
summary(aov_model)
```

Group Means

```{r}
pedon_df %>%
  group_by(order) %>%
  summarise(
    mean_carbon = mean(total_c_1m, na.rm = TRUE),
    sd_carbon = sd(total_c_1m, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```

5. Regression: Carbon vs Latitude

```{r}
pedon_df_lat <- pedon_df %>%
  filter(lat > 40, lat < 65)

lm_model <- lm(total_c_1m ~ lat, data = pedon_df_lat)
summary(lm_model)
```

Plot Regression
```{r}
# 5. Regression: Carbon vs Latitude + Soil Order (Improved Model)

# 先清理数据（过滤异常纬度）
pedon_df_mod <- pedon_df %>%
  filter(lat > 40, lat < 65)

# 建立多元回归模型（关键升级）
lm_model2 <- lm(total_c_1m ~ lat + order_group, data = pedon_df_mod)

# 查看结果
summary(lm_model2)

# 更清晰的系数表（可用于报告）
tidy(lm_model2)

# 按 soil group 分开看趋势（强烈推荐）
ggplot(pedon_df_mod, aes(x = lat, y = total_c_1m, color = order_group)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ order_group) +
  theme_classic() +
  theme(legend.position = "none")+
  labs(subtitle = "Slopes vary across soil groups, indicating weak but heterogeneous relationships")

lm_model3 <- lm(total_c_1m ~ lat * order_group, data = pedon_df_mod)

summary(lm_model3)
```


Horizon Contribution in Mineral Soils

```{r}
mineral_horizon_summary <- df %>%
  mutate(
    order_group = case_when(
      str_detect(order, regex("Histosol|Folisol|Mesisol|Humisol|Cryo", ignore_case = TRUE)) ~ "Organic soils",
      str_detect(order, regex("Podzol", ignore_case = TRUE)) ~ "Podzol",
      str_detect(order, regex("Brunisol|Brunisolic", ignore_case = TRUE)) ~ "Brunisol",
      str_detect(order, regex("Gleysol|Gleysolic", ignore_case = TRUE)) ~ "Gleysol",
      str_detect(order, regex("Regosol|Regosolic", ignore_case = TRUE)) ~ "Regosol",
      TRUE ~ "Other"
    )
  ) %>%
  filter(order_group != "Organic soils") %>%
  filter(!is.na(horizon_type)) %>%
  group_by(horizon_type) %>%
  summarise(
    total_carbon = sum(ccontent_1m, na.rm = TRUE),
    mean_carbon = mean(ccontent_1m, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(prop_carbon = total_carbon / sum(total_carbon))

mineral_horizon_summary
```


7. Notes

* Analysis is conducted at the pedon level (independent observations)
* Soil order is used instead of horizon-based classification
* ANOVA tests group differences
* Regression explores geographic trends
