---
title: "Final_project_analysis_v3"
output: html_document
---


```{r setup, message=FALSE, warning=FALSE}

library(tidyverse)

library(broom)
```


1. Load Data

```{r}
df <- read_csv("../data/data_final_project.csv")

glimpse(df)
```

2. Create Pedon-level Dataset

```{r}
pedon_df <- df %>%
  filter(pedon_start == TRUE) %>%
  mutate(
    order_group = case_when(
      str_detect(order, regex("Histosol|Folisol|Mesisol|Humisol|Cryo", ignore_case = TRUE)) ~ "Organic soils",
      str_detect(order, regex("Podzol", ignore_case = TRUE)) ~ "Podzol",
      TRUE ~ "Other"
    )
  ) %>%
  select(pedon_id, order, order_group, total_c_1m, lat) %>%
  filter(!is.na(total_c_1m), !is.na(order_group))

pedon_df <- pedon_df %>%

  mutate(

    order_group = factor(

      order_group,

      levels = c("Organic soils", "Podzol", "Other")

    )

  )
# check structure
glimpse(pedon_df)

# check sample sizes
pedon_df %>% count(order_group)
```

3. Exploratory Analysis

Boxplot: Soil Carbon by Soil Group

```{r}
ggplot(pedon_df, aes(x = order_group, y = total_c_1m, fill = order_group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.1) +
  theme_classic() +
  labs(
    title = "Soil Carbon by Soil Group",
    x = "Soil Group",
    y = "Total Carbon (0–1 m)"
  )
```


4. ANOVA: Does Carbon Differ Across Soil Groups?
```{r}
aov_model <- aov(total_c_1m ~ order_group, data = pedon_df)
summary(aov_model)
```

Group Median & Means

```{r}
pedon_df %>%
  group_by(order_group) %>%
  summarise(
    mean_carbon = mean(total_c_1m, na.rm = TRUE),
    median_carbon = median(total_c_1m, na.rm = TRUE),
    sd_carbon = sd(total_c_1m, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```

5. Regression: Carbon vs Latitude

Filter realistic latitude range
```{r}

pedon_df_lat <- pedon_df %>%
  filter(lat > 40, lat < 65)
```

Scatter plot + regression line

```{r}
ggplot(pedon_df_lat, aes(x = lat, y = total_c_1m)) +
  geom_point(alpha = 0.2, size = 1.5) +
  geom_smooth(method = "lm", color = "blue", linewidth = 1.2, se = TRUE) +
  theme_classic(base_size = 13) +
  labs(
    title = "Soil Carbon vs Latitude",
    x = "Latitude",
    y = "Total Carbon (0–1 m)"
  )
```

Simple Linear Regression
```{r}
lm_model <- lm(total_c_1m ~ lat, data = pedon_df_lat)
summary(lm_model)
```

6. Improved Model: Add Soil Group
```{r}
lm_model2 <- lm(total_c_1m ~ lat + order_group, data = pedon_df_lat)

summary(lm_model2)

tidy(lm_model2)
```

Faceted visualization (by soil group)
```{r}
ggplot(pedon_df_lat, aes(x = lat, y = total_c_1m, color = order_group)) +
  geom_point(alpha = 0.15, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  facet_wrap(~ order_group) +
  scale_x_continuous(
    limits = c(48, 60),
    breaks = seq(48, 60, by = 3)
  ) +
  theme_classic(base_size = 13) +
  labs(
    title = "Soil Carbon vs Latitude by Soil Group",
    x = "Latitude",
    y = "Total Carbon (0–1 m)"
  )
```

7. Interaction Model
```{r}
lm_model3 <- lm(total_c_1m ~ lat * order_group, data = pedon_df_lat)

summary(lm_model3)
```


8. Horizon Contribution in Mineral Soils
```{r}
mineral_horizon_summary <- df %>%
  mutate(
    order_group = case_when(
      str_detect(order, regex("Histosol|Folisol|Mesisol|Humisol|Cryo", ignore_case = TRUE)) ~ "Organic soils",
      str_detect(order, regex("Podzol", ignore_case = TRUE)) ~ "Podzol",
      TRUE ~ "Other"
    )
  ) %>%
  filter(order_group != "Organic soils") %>%
  filter(!is.na(horizon_type)) %>%
  group_by(horizon_type) %>%
  summarise(
    total_carbon = sum(ccontent_1m, na.rm = TRUE),
    mean_carbon = mean(ccontent_1m, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(prop_carbon = total_carbon / sum(total_carbon))

mineral_horizon_summary
```

9. Notes

* Analysis conducted at pedon level (independent observations)
* Soil classification simplified to improve interpretability
* ANOVA used to test group differences
* Regression used to explore spatial trends
* Horizon analysis examines internal carbon distribution
